---
title: "combine_halk"
output: html_document
date: "2025-09-04"
editor_options: 
  chunk_output_type: console
---

# Library
```{r}
library(tidyverse)
library(arrow)
```

# Read in data
The data in the HALKed files are the lengthen and lengthened fish of the 8 sport fish species, we really only care about the fish that go an age here - so lets filter for where an alk was applied
```{r}
#read in state specific halked fish data filepath
ia_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  filter(!is.na(alk))

#read in state specific halked fish data filepath
in_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  filter(!is.na(alk))

#read in state specific halked fish data filepath
il_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  filter(!is.na(alk))

#read in state specific halked fish data filepath
mi_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  filter(!is.na(alk))

#read in state specific halked fish data filepath
sd_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  filter(!is.na(alk)) %>% 
  mutate(state = "South_Dakota")
#small data join
small <- bind_rows(ia_data, il_data, mi_data, sd_data, in_data)
rm(ia_data, il_data, mi_data, sd_data, in_data)

#read in state specific halked fish data filepath
wi_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  filter(!is.na(alk))
small2 <- bind_rows(small, wi_data)
rm(wi_data, small)

#read in state specific halked fish data filepath
mn_data <- open_dataset(sources = file.path()) %>% 
  collect() %>% 
  mutate(obs.id = as.character(obs.id)) %>% 
  filter(!is.na(alk))

mw_data <- bind_rows(small2, mn_data)
rm(small2, mn_data)
```

# adding a filter for length of fish that get assigned an age
This filtering is a step to make sure that fish that get assigned an age are within the length range of the given creation of the ALK. Example, there may only be age3+ walleye aged in the creation of an alk, we need to make sure that a super small walleye (~100mm, age-0) that would get assigned age-3 are not included 
```{r}
#we will need HUC and ecoregion data for this
#read in huc and ecoregion data 
load()
load()

mw_data <- mw_data %>% 
  mutate(state = case_when(state == "South_Dakota" ~ "South Dakota",
                           TRUE ~ state)) %>% 
  left_join(
        lake_huc10 %>% 
          ungroup() %>% 
          mutate(lake.id = gsub("^0", "", lake.id)) %>% 
          left_join(
            data.frame(state = tolower(state.abb), state.name),
            by = "state"
          ) %>%  select(-state) %>%  rename(state = state.name),
        by = c("state", "lake.id")
      ) %>% 
      left_join(
        ecoregions %>% 
          mutate(lake.id = gsub("^0", "", lake.id)) %>% 
          left_join(
            data.frame(state = tolower(state.abb), state.name),
            by = "state"
          ) %>%  
          select(-state) %>% 
          rename(state = state.name),
        by = c("state", "lake.id")
      ) %>% 
      rename(huc10.code = huc.code) %>% 
      mutate(huc8.code = substr(huc10.code, 1, 8)) %>% 
  mutate(aging.structure.filter = case_when(str_detect(aging.structure, "tolith") ~ "otolith",
                                            str_detect(aging.structure, "pine") ~ "spine",
                                            TRUE ~ aging.structure)) %>%
  #all fish in cm for filtering
  mutate(filtering_length = case_when(length.unit == "mm" ~ length/10,
                                      length.unit == "in" ~ length*2.54,
                                      TRUE ~ NA))

mw_data %>% 
  summarise(problem_huc = sum(is.na(huc10.code) & alk == "huc10.code"),
            problem_eco = sum(is.na(level3_ecoregion) & alk == "level3_ecoregion"))
#our joins work, all alks at the huc/eco level have that info now
rm(ecoregions,
   lake_huc10)

#do our aging structures look alright?
mw_data %>% 
  filter(!is.na(age)) %>% 
  group_by(aging.structure, 
           alk.age.str,
           aging.structure.filter) %>% 
  count() %>% 
  print(n = nrow(.))

#lets make a function that will group the correct fish based on the alk and calcuate a length buffer
make_cutoffs <- function(data, group_vars, level_name, buffer_prop = 0.05) {
  # Only drop NAs for the grouping vars - this makes sure NA hucs are removed when HUC is important but we can still get a cutdown at the ecoregion level 
  data_filtered <- data %>%
    filter(!is.na(age)) %>%
    filter(across(all_of(group_vars), ~ !is.na(.x))) %>% 
    #filter only for fish that were used in the alk creation (aging structure of alk)
    filter(aging.structure.filter == alk.age.str)
  
  data_filtered %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      alk_min_length = min(filtering_length, na.rm = TRUE),
      alk_max_length = max(filtering_length, na.rm = TRUE),
      # keep other hierarchy columns as-is
      huc8.code        = first(huc8.code),
      huc10.code       = first(huc10.code),
      level3_ecoregion = first(level3_ecoregion),
      state = first(state),
      .groups = "drop"
    ) %>%
    mutate(
      buffer  = buffer_prop * (alk_max_length - alk_min_length),
      low_cut  = alk_min_length - buffer,
      high_cut = alk_max_length + buffer,
      alk = level_name
    )
}
  
cut_lakeyear <- make_cutoffs(mw_data, c("lake.id","year","species", "state"), "year")
cut_lake     <- make_cutoffs(mw_data, c("lake.id","species", "state"), "lake.id") 
cut_huc10    <- make_cutoffs(mw_data, c("huc10.code","species"), "huc10.code") %>% 
  select(-huc8.code, -level3_ecoregion, -state)
cut_huc8     <- make_cutoffs(mw_data, c("huc8.code","species"), "huc8.code") %>% 
  select(-huc10.code, -level3_ecoregion, -state)
cut_ecoreg   <- make_cutoffs(mw_data, c("level3_ecoregion","species"), "level3_ecoregion") %>% 
  select(-huc10.code, -huc8.code, -state)
cut_species  <- make_cutoffs(mw_data, c("species"), "species") %>% 
  select(-huc10.code, -huc8.code, -level3_ecoregion, -state)

join_lakeyear <- mw_data %>% 
  filter(alk == "year") %>% 
  left_join(cut_lakeyear)
join_lake <- mw_data %>% 
  filter(alk == "lake.id") %>% 
  left_join(cut_lake)
join_huc10 <- mw_data %>% 
  filter(alk == "huc10.code") %>% 
  left_join(cut_huc10)
join_huc8 <- mw_data %>% 
  filter(alk == "huc8.code") %>% 
  left_join(cut_huc8)
join_eco <- mw_data %>% 
  filter(alk == "level3_ecoregion") %>% 
  left_join(cut_ecoreg)
join_spec <- mw_data %>% 
  filter(alk == "species") %>% 
  left_join(cut_species)

mw_data_with_cuts <- bind_rows(join_eco,
                               join_huc10,
                               join_huc8,
                               join_lake,
                               join_lakeyear,
                               join_spec)
rm(join_eco,
   join_huc10,
   join_huc8,
   join_lake,
   join_lakeyear,
   join_spec)

rm(cut_lakeyear,
   cut_lake,
   cut_huc10,
   cut_huc8,
   cut_ecoreg,
   cut_species)

glimpse(mw_data_with_cuts)

#just a few check that I am able to group by the correct 
mw_data_with_cuts %>% 
  filter(lake.id == "11014700" & species == "walleye" & aging.structure.filter == alk.age.str & year == 2010) %>% 
  summarise(n = n())

mw_data_with_cuts %>% 
  filter(lake.id == "11014700" & species == "walleye" & year == 2010) %>% 
  distinct(alk.n)
#that is a good gut check

mw_data_with_cuts %>% 
  filter(species == "walleye" & aging.structure.filter == "spine" & level3_ecoregion == "driftless_area" & !is.na(age)) %>% 
  count()

mw_data_with_cuts %>% 
  filter(alk == "level3_ecoregion" & species == "walleye" & level3_ecoregion == "driftless_area") %>% 
  distinct(alk.n,
           alk.age.str)

#when i make the actual cuts, it will need to be with the filtering length 

mw_data_with_cuts <- mw_data_with_cuts %>% 
  filter(filtering_length > low_cut & filtering_length < high_cut)

diff <- mw_data %>% 
  anti_join(mw_data_with_cuts)

diff %>% 
  group_by(is.na(age)) %>% 
  count()

diff %>% 
  group_by(state) %>% 
  count()

diff %>% 
  summarise(has_ages = sum(!is.na(age)),
            matches_str = sum(!is.na(age) & aging.structure.filter == alk.age.str, na.rm = T),
            has_est_age = sum(!is.na(est.age)),
            n = n())
# we can see here that all cut records have an estimated age. The fish that have been cut with an age is due to a different aging structure used in the halk

rm(diff, mw_data, make_cutoffs)
```


# QA/QC
```{r}
names(mw_data_with_cuts)

mw_data_with_cuts %>% 
  group_by(state) %>% 
  summarise(length_bin = sum(!is.na(length.bin)),
            lengths = sum(!is.na(length)),
            length_bin_calc = sum(!is.na(length) & !is.na(length.bin)),
            n = n()) 

mw_data_with_cuts %>% 
  group_by(length.unit, state) %>% 
  count()

mw_data_with_cuts %>% 
  group_by(species, length.unit, state) %>% 
  summarise(min = min(length),
            med = median(length),
            max = max(length)) %>% 
  print(n = nrow(.))
#ages get assigned to some pretty crazy lengths, but we cant control what the agency reports

mw_data_with_cuts %>% 
  filter(state == "South Dakota" & species == "walleye" & length > 5000) %>% 
  glimpse()

mw_data_with_cuts %>% 
  group_by(species) %>% 
  count()

mw_data_with_cuts %>% 
  group_by(state) %>% 
  summarise(n_obs = n_distinct(obs.id),
            n = n()) %>% 
  mutate(test = n_obs == n)

check <- mw_data_with_cuts %>% 
  filter(!is.na(length.bin)) %>% 
  select(length, length.bin, state) 
rm(check)

mw_data_with_cuts %>% 
  group_by(state) %>% 
  summarise(aged_fish = sum(!is.na(age) & !is.na(est.age)))

mw_data_with_cuts %>% 
  group_by(state, alk) %>% 
  count() %>% 
  print(n = nrow(.))

mw_data_with_cuts %>% 
  group_by(alk.age.str) %>% 
  count() 

#the below lakes were examples of not having huc information 
#BEA25 did not have huc but should return eco
#FRM00 did not have either, and should return species 
#prior to fixing NAs when assigning ages, they were returning huc10
mw_data_with_cuts %>% 
  filter(lake.id %in% c("BEA25", "FRM00")) %>% 
  filter(species == "walleye") %>% 
  distinct(lake.id, obs.id, est.age, alk)

#here is our mn example, this should return ecoregion 
mw_data_with_cuts %>% 
  filter(lake.id == "86017800") %>% 
  distinct(lake.id, alk)

mw_data_with_cuts %>% 
  group_by(state) %>% 
  count()

mw_data_with_cuts %>% 
  summarise(min = min(length, na.rm = T),
            max = max(length, na.rm = T)) %>% 
  collect()

mw_data_with_cuts %>% 
  filter(length == 0) %>% 
  group_by(is.na(length.bin)) %>% 
  count() %>% 
  collect()

mw_data_with_cuts %>% 
  filter(state == "Iowa" & length.bin == "0-9.99") %>% 
  group_by(length) %>% 
  count() %>% 
  collect()
```

# making cross walk
```{r}
mw_data <- mw_data_with_cuts %>% 
  rename(alk_length = length,
         obs_id = obs.id,
         est_age = est.age,
         alk_age_str = alk.age.str,
         alk_n = alk.n) %>% 
  select(state,
         obs_id,
         est_age,
         alk,
         alk_age_str,
         alk_length,
         alk_n)
#write_csv(mw_data, ".csv")
```

# check winni example  
```{r}
check <- mn_data %>% 
  filter(lake.id == "11014700" & year == "2010")

check %>% 
  filter(species == "walleye" & !is.na(age)) %>% 
  ggplot() +
  geom_point(aes(age, length, color = aging.structure, shape = sampling.method))

check %>% 
  filter(species == "walleye") %>% 
  ggplot() +
  geom_point(aes(est.age, length, color = alk.age.str))
```

